---
title: "j0j0r vignette"
output:
  prettydoc::html_pretty:
    theme: tactile
    highlight: github
    math: katex
vignette: >
  %\VignetteIndexEntry{j0j0r vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(magrittr)
library(rlang)
library(j0j0r)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The j0j0r package is meant to calculate the unscreened current correlation tensor, $\langle \widetilde{\bf{j}}^{(a0)} \widetilde{\bf{j}}^{(a0)}\rangle$, of the plasma fluctuation model of [Bindslev 1996, Journal of Atmospheric and Terrestial Physics, **58**, 983](https://www.sciencedirect.com/science/article/pii/0021916995001298). 

The elements of $\langle \widetilde{\bf{j}}^{(a0)} \widetilde{\bf{j}}^{(a0)}\rangle$ are given by Eq. 25 and 26 of [Bindslev 1996](https://www.sciencedirect.com/science/article/pii/0021916995001298) or by Eq. 7 and 8 of [Stejner et al 2011 PPFE, **53**, 065020](https://orbit.dtu.dk/files/5555263/Stejner_preprint.pdf):

$$\langle \widetilde{\bf{j}}^{(a0)} \widetilde{\bf{j}}^{(a0)}\rangle = (2\pi)^2 \frac{m_a q_a^2}{\lvert k_\parallel \rvert} \int dp_\perp p_\perp \sum_{l = -\infty}^\infty {\bf c}_l {\bf c}_l^*f^{(a0)}(p_\perp, p_\parallel)$$

and

$$ {\bf c}_l = \left(\begin{array}{c} \frac{l\omega_{ca}}{k_\parallel}J_l(k_\perp\rho) \\ -iv_\perp J'_l(k_\perp\rho) \\ v_\parallel J_l(k_\perp\rho) \end{array}\right), \quad p_\parallel = m_a\frac{\omega -l\omega_{ca}}{k_\parallel}, \quad \rho = v_\perp/\omega_{ca}, \quad \omega_{ca} = q_aB^{(0)}/m_a $$

with notation as in the references. 

The j0j0r code is meant to be distribution-agnostic, i.e. it should be able to handle any reasonable form of the momentum distribution. The integral over perpendicular momentum is calculated numerically with [stats::integrate](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/integrate.html). Analytic solutions, possible only in special cases, are not used/supported. 

The code is therefore meant to be flexible and easy to use, but it is not designed for speed. However, the code does support parallelized calculations using the [future](https://cran.r-project.org/web/packages/future/vignettes/future-1-overview.html) and [furrr](https://cran.r-project.org/web/packages/furrr/index.html) package.

# Distribution functions

New momentum distributions can be input in a specific format (described below). They do not need to be included in the package, but at present the package includes functions to set up the following distribution types:

* A simple isotropic Maxwellian distribution, see `maxwellian_setup()`. 

* A bi-Maxwellian distribution, with drift along the magnetic field (z-direction), see `j0j0r::bimaxwellian_setup()`.

* A generalized Lorentzian / Kappa distribution, see [here](https://www.spenvis.oma.be/help/background/distributions/distributions.html) and `j0j0r::generalized_lorentzian_setup()`.

* A "Maxwellian-like" ring-distribution, see `j0j0r::maxwellian_ring_setup()`.

* A bivariate normal distribution, see `j0j0r::bvtnorm_setup()`.

* A "Wilkie" distribution: an isotropic fast-ion slowdown with transport modifications as formulated by George J. Wilkie,  see [https://arxiv.org/abs/1808.01934v2](https://arxiv.org/abs/1808.01934v2) and `j0j0r::wilkie_setup()`.  

### Conventions

A particle with specified momentum distribution can be input to the code as a list with elements:

* name, a text string: some name for the particle, e.g. "Deutrium".
* A, an integer: ion mass number.
* Z, an integer: ion charge number.
* distribution, a list with elements:
    * function_name, a text string:  name of function to be called to evaluate the distribution.
    * gradient, a text string:  name of function to be called to be called to evaluate the gradient of the distribution. Not used at present, so optional. 
    * distargs, a list of name-value pairs: all arguments to the distribution function other than p_par and p_perp, e.g. density and thermal momentum.
    * p_scale, a double: a typical/representative momentum scale. Used to scale momentum so it is near 1 in calls to `stats::integrate()`.

Such a list does not have to be constructed using functions from the package, but the package includes some examples (mentioned above) of how this can be done. 

The code assumes that the distributions are in cylindrical coordinates and symmetric about the magnetic field, so ${\bf p = (p_\perp, p_\parallel)$ where the magnetic field is in the z-direction. Thus, $p_\perp =\sqrt{(p_x^2 + p_y^2)}$ and $p_\parallel = p_z$. When transforming between different coordinate systems the distribution function must remain normalized, so

$$ 
n = \int f({\bf p}) d{\bf p} = \int f(p_x, p_y, p_z)dp_x dp_y dp_z = 2\pi\int f(p_\perp, p_\parallel) p_\perp dp_\perp dp_\parallel 
$$
Where the final equality again assumes symmetry around the megnetic field (z-axis), so the integral over azimuthal angle contributes a factor $2\pi$. 

Note that with this convention, the factor $2\pi p_\perp$ is considered part of the volume element, not of the distribution function $f(p_\perp, p_\parallel)$. So with this convention the expression for a Maxwellian velocity distribution is 

$$
f_{Mw}(p_\perp, p_\parallel) = \frac{n}{(\sqrt{\pi}p_t)^3}e^{-(p_\perp^2 + p_\parallel^2)/p_t}, \qquad
p_t = \sqrt{2k_bTm}
$$
where $T$ is temperature, $k_b$ Boltzmann's constant and $p_t$ is the thermal momentum corresponding to the thermal velocity $v_t = \sqrt{2k_bT/m}$.

### Examples
A Maxwellian distribution and a Wilkie slowdown distribution can be set up with:

```{r, echo = TRUE, results = FALSE}
maxwellian_deuterium <- j0j0r::maxwellian_setup(
  n = 4e19,
  T_eV = 2000,
  A = 2,
  Z = 1,
  name = "maxwellian"
)

wilkie_deuterium <- j0j0r::wilkie_setup(
  b = 3,
  n = 4e19,
  A = 2,
  Z = 1,
  birth_energy = 10e3,
  n_e = 4e19,
  T_e_eV = 2e3,
  ions = data.frame(
    Z = 1,
    A = 2,
    n = 4e19
  ),
  name = "wilkie"
)
```

And they can be plotted as a function of parallel velocity with 

```{r fig1, echo = TRUE, fig.height = 6, fig.width = 6, fig.align = "center"}
j0j0r::calculate_distribution_data_frame(
  particles = list(
    maxwellian = maxwellian_deuterium, 
    wilkie = wilkie_deuterium
    ), 
  v_par = seq(-1.5e6, 1.5e6, length.out = 1000), 
  v_perp = 0
) %>% 
  ggplot2::ggplot(
    mapping = ggplot2::aes(x = v_par, y = value, color = name)
    ) +
  ggplot2::geom_line(size = 1.2) +
  ggplot2::theme(legend.position = "top") +
  ggplot2::ylab("Density") +
  ggplot2::xlab(latex2exp::TeX("$v_{par}$ in m/s^2")) +
  ggplot2::theme(text = ggplot2::element_text(size = 17))
```

The `j0j0r::calculate_distribution_data_frame()` function returns a data frame wit the evaluated distribution for all combinations of its arguments: particles, v_par and v_perp. So a 2D-plot can be obtained by allowing both v_par and v_perp to vary:

```{r fig2, echo = TRUE}
j0j0r::calculate_distribution_data_frame(
  particles = list(
    maxwellian = maxwellian_deuterium, 
    wilkie = wilkie_deuterium
    ), 
  v_par = seq(-1e6, 1e6, length.out = 300), 
  v_perp = seq(0, 1e6, length.out = 300)
) %>% 
  ggplot2::ggplot(
    mapping = ggplot2::aes(y = v_par, x = v_perp, z = value)
    ) +
  ggplot2::geom_raster(ggplot2::aes(fill = value)) +
  ggplot2::geom_contour(colour = "white", alpha = 0.2) +
  ggplot2::facet_wrap( ~ name) +
  viridis::scale_fill_viridis(option = "plasma", limits = c(0, 3e81)) +
  ggplot2::ylab(unname(latex2exp::TeX("$v_{par}$ in m/s^2"))) +
  ggplot2::xlab(latex2exp::TeX("$v_{perp}$ in m/s^2")) +
  ggplot2::guides(fill = ggplot2::guide_colourbar(title = "Density")) +
  ggplot2::theme(
    text = ggplot2::element_text(size = 15),
    axis.text.x = ggplot2::element_text(angle = -45)
    ) +
  ggplot2::scale_x_continuous(labels = scales::scientific)
```

The normalization can be checked with the function `j0j0r::integrate_distribution()`. For example:

```{r, echo = TRUE, results = asis}
j0j0r::integrate_distribution(distribution = maxwellian_deuterium$distribution)
```


# Calculating j0j0

The function `j0j0r::j0j0_element()` is used to calculate a single element of j0j0 for a single wave vector, frequency, distribution etc. The function `j0j0r::j0j0()` is a wrapper for parameter sweeps. It requires the following input:

* k: length of fluctuation wave vector. No dispersion function is yet included with the j0j0r package, so below k is simply set to $2\pi  / \lambda$ with $\lambda = c/f = c/100e9\,s^{-1} = 0.003\,m$ 

* phi: angle (in degrees) between wave vector and magnetict field, $\phi = \angle({\bf k}, {\bf B})$.

* frequencies: fluctuation frequency in Hz.

* direction: one or more spatial directions ("x", "y" or "z").

* B strength of magnetic field in Tesla.

* particle: a list with with mass, charge, and momentum distribution. As in the examples above.

All of these inputs can ved vectors or lists with multiple elements, e.g. multiple frequencirs or particles. The function will form all unique combinations of its inputs and return $\langle \widetilde{\bf{j}}^{(a0)} \widetilde{\bf{j}}^{(a0)}\rangle$ for each. So a sweep of frequency and direction for two values of phi can be done with:

```{r, echo = TRUE, results = FALSE}
specdf <- j0j0r::j0j0(
  k = 2 * pi / (j0j0r::const$c / 100e9),
  phi = c(60, 87),
  frequencies = seq(0, 400e6, by = 1e6),
  directions = c("x", "y", "z"),
  B = 2.5,
  particles = list(
    maxwellian = maxwellian_deuterium
  )
)
```

And plotted with:

```{r, echo = TRUE, results = FALSE}
specdf %>% 
  ggplot2::ggplot(
    mapping = ggplot2::aes(
      x = frequency/1e6, 
      y = Re(j0j0) + Im(j0j0), 
      color = forcats::fct_rev(as.factor(phi))
      )
    ) +
  ggplot2::geom_line(size = 1) +
  ggplot2::geom_hline(yintercept = 0) +
  ggplot2::facet_wrap( ~ directions) +
  ggplot2::xlab("Frequency in MHz") +
  ggplot2::guides(color = ggplot2::guide_legend(title = expression(phi))) + 
  ggplot2::theme(legend.position = "top",
                 text = ggplot2::element_text(size = 13)
                 )
  

  
```



# Examples
